---
title: "Final_Project"
format: html
editor: visual
---

# COVID Outcomes in Tennessee 2020-2024

## Load packages and download files

```{r}
library(tidyverse)
library(readr)
library(ggplot2)
# Open CSV File 
COVID19_Outcomes_Tennessee <- read_csv("../data/COVID19_Outcomes_Tennessee_by_Race_Sex.csv")
```

## Tidy Data

Convert Date to proper format

```{r}
COVID19_Outcomes_Tennessee <- COVID19_Outcomes_Tennessee |>
  mutate(
    WEEK_END_DATE = as.Date(WEEK_END_DATE, format = "%Y %b %d %I:%M:%S %p"))
COVID19_Outcomes_Tennessee
```

Clean column names and pivot longer

```{r}
COVID19_Outcomes_clean <- COVID19_Outcomes_Tennessee |>
  rename(
    group = CAT_DETAIL,
    date = WEEK_END_DATE,
    cases = CAT_TOTALCASES,
    deaths = CAT_TOTALDEATHS,
    hospitalizations = CAT_TOTALHOSP,
    category = CATEGORY
  )

COVID19_Outcomes_clean <- COVID19_Outcomes_clean |>
  pivot_longer(
    cols = c(cases, deaths, hospitalizations),
    names_to = "outcomes",
    values_to = "count"
  )
COVID19_Outcomes_clean
```

# Questions

## Which racial group had the highest total COVID-19 cases per population?

Create a table with race only

```{r}
race_data <- COVID19_Outcomes_clean |>
  filter(category == "RACE",outcomes == "cases" ) |> filter(date == max(date))
race_data
```

Open population table and tidy

```{r}
population_2022 <- read_csv("../data/population_2022.csv")
population_2022 <- population_2022 |>  slice(c(39, 40, 41, 46, 54, 60)) |>
  select("Label (Grouping)", "Tennessee!!Estimate") |>
  rename( group = "Label (Grouping)", 
          population = "Tennessee!!Estimate") |>
  mutate(
    group = str_squish(group),
    population = as.numeric(population)
  ) 
population_2022 <- population_2022  |>  mutate(
    group = case_when(
      group == "White" ~ "White",
      group == "Black or African American" ~ "Black or African American",
      group == "American Indian and Alaska Native" ~ "American Indian or Alaska Native",
      group == "Asian" ~ "Asian",
      group == "Native Hawaiian and Other Pacific Islander" ~ "Native Hawaiian or Other Pacific Islander",
      group == "Two or More Races" ~ "Other/Multiracial"
    )
  ) 
population_2022
```

Join the data and arrange in descending order

```{r}
race_joined <- race_data |> 
  left_join(population_2022, by = "group") |> 
  mutate(case_rate_per_100k = (count / population) * 100000) |>
  arrange(desc(case_rate_per_100k))
race_joined
```

## Did COVID-19 death rates differ by gender relative to case counts?

Filter for gender only

```{r}
gender_data <- COVID19_Outcomes_clean |>
  filter(category == "SEX",
         outcomes %in% c("cases", "deaths"))
gender_data
```

Get most recent gender data

```{r}
gender_latest <- gender_data |>
  filter(date == max(date)) 
gender_latest

```

Pivot wider so cases and death are own category and calculate

```{r}
gender_wide <- gender_latest |>
  pivot_wider(
    names_from = outcomes,
    values_from = count
  )
gender_rates <- gender_wide |>
  mutate(
    death_rate = deaths / cases
  )
gender_rates <- gender_rates |>
  mutate(
    death_rate_percent = death_rate * 100
  )
```

Create a visualization

```{r}
ggplot(gender_rates, aes(x = group, y = death_rate_percent)) +
  geom_col() +
  labs(
    title = "COVID-19 Death Rate by Gender",
    x = "Gender",
    y = "Death Rate (%)"
  )
```

## How did COVID-19 case rates change over time?

Filter the data set to only focus on race and case data

```{r}
race_data <- COVID19_Outcomes_clean |>
  filter(category == "RACE", outcomes == "cases")
race_data
```

Summarize cases by race and date. We combine repeated entries so each race has one total case value per week

```{r}
race_summary <- race_data |>
  group_by(group, date) |>
  summarise(
    cases = sum(count, na.rm = TRUE),
    .groups = "drop"
  )
race_summary
```

Create a time trend line graph

```{r}
plot <- ggplot(race_summary, aes(x = date, y = cases, color = group, group = group)) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  labs(
    title = "COVID-19 Case Trends Over Time by Race in Tennessee",
    x = "Date",
    y = "Cases",
    color = "Race"
  )
plot
```

Save figure to directory

```{r}
ggsave("../figures/covid_race_trends.png", plot = plot, width = 8, height = 5)
```
